Towards a Practical Face Recognition System: \\ Robust Alignment and Illumination by Sparse Representation

### 2011-01-07 faces_system.tex
Some language fixes from Drew.

\PARstart{F}{ace} recognition applications to date have fallen into roughly two
categories.  Face recognition has recently seen a lot of success in a family of
less-demanding applications such as online image search and family photo album
organization (e.g.\ Google Picassa, Microsoft Photo Gallery, and Apple iPhoto).
At the other end of the tractability spectrum there are the terrorist watchlist
and mass surveillance applications that have for the most part dominated the
field of face recognition research.  However, there are many face recognition
that fall roughly between these extremes, where very high recognition
performance is desired, but the users in the gallery are still allies of the
system rather than adversaries.  These applications include access control for
secure facilities (e.g., prisons, office buildings), computer systems,
automobiles, or automatic teller machines, where controlled gallery images can
be obtained in advance.  These applications are very interesting due to their
potential sociological impact.  Since the gallery subjects are allies, rather
than opponents, of the recognition system, this creates the possibility of
carefully controlling the acquisition of the training data. While the same can
be said for other biometrics such as fingerprints and iris recognition, face
recognition has the potential of working with test data that is much less
controlled, allowing the access control system to be made less intrusive to the
users of the system. To some extent, the goal of this paper is to show how a
reliable face recognition system can be built for this admittedly restricted,
but still important, scenario.

@@

Unfortunately, while the work \cite{Wright2009-PAMI} achieves impressive results, it does not deal with misalignment
between the test and training images, and it requires a rich set of illuminations
in the gallery images for good performance.  
%Furthermore, the experiments were carried out on public datasets taken under
%controlled laboratory conditions such as Extended Yale B
%\cite{Georghiades2001-PAMI}, which do not contain realistic test images.
%two critical aspects of real world face recognition must be addressed:  significant variations in both the {\em image domain} and in
%the {\em image value}.
We illustrate the compounded effect of both alignment and illumination 
with an example in Figure \ref{fig:promo}.
The task is to identify the girl among 20 subjects. If the test
face image, say obtained from an off-the-shelf face detector,
has even a small amount of registration error against the
training images (caused by mild pose, scale, or misalignment),
the sparse representation obtained using the method of \cite{Wright2009-PAMI} is no longer informative, 
even if sufficient illuminations are present in the training, as shown in Figure
\ref{fig:promo}(top). In addition, in order to span
the illuminations of a typical indoor (or outdoor) environment,
illuminations from behind the subject are also needed in the
training. Otherwise, even for perfectly aligned test images,
the sparse representation obtained using \cite{Wright2009-PAMI} will not necessarily be sparse or
informative, as shown by the example in Figure
\ref{fig:promo}(middle). Clearly, both good alignment, as well
as sufficient training images are needed to ensure success of the sparsity-based 
recognition method proposed by \cite{Wright2009-PAMI}.  
%Unfortunately, most public face databases lack images with a
%significant component of rear (more than 90 degrees from
%frontal) illumination, either for training or testing.
In this paper, we examine how to handle alignment and illumination simultaneously
in the sparse representation framework, hence bring the method proposed in \cite{Wright2009-PAMI}
closer to practical use.

@@

We briefly review existing techniques for recognition, image registration, and
handling of illumination variation.  Our system is based purely on 2D
techniques.  This fact immediately distinguishes our approach from systems
that either require a 3D data as an input, or attempt to estimate a 3D model
from 2D input \cite{amberg2007reconstructing,Blanz2003-PAMI}.  While these
techniques can achieve better robustness to pose variation given
a sufficiently accurate 3D model, for access control applications where only 
moderate pose variation is present, the proposed method will be more
than sufficient.   Note that according to the principle of photometric stereo, images of faces
under varying illuminations already contain 3D shape-related information,\footnote{In principle,
one can recover the 3D shape of the face from multiple illuminations using photometric stereo.} 
and this information can be leveraged by 2D algorithms for alignment and recognition 
even if shape is not reconstructed explicitly. In addition, the proposed sparsity-based robust
alignment and recognition techniques are not limited to 2D data and can be easily extended
to 3D data if necessary.

% Alignment
In holistic recognition algorithms, correspondence between points in the test
image and in the training must be achieved.  A long line of research exists on
using Active Appearance Models \cite{Cootes2001-PAMI}, and the closely related
Active Shape Models \cite{cootes1992active} to register images against a
relatively high-dimensional model of plausible face appearances, often
leveraging face-specific contours.  While these model-based techniques have
advantages in dealing with variations in expression and pose, they may add
unnecessary complexity to applications where subjects normally present a neutral face
or only have moderate expression. We prefer to focus on deformations with far fewer 
degrees of freedom (say an affine transform), and to use the training images themselves 
as the appearance model. Iterative registration in this spirit dates at least back to the 
Lucas-Kanade algorithm \cite{lucas1981iterative}.

However, whereas much of the early work on image registration is aimed at the
problem of registering nearly identical images, say by minimizing a sum of
squared distances or maximizing normalized correlation, here we must confront several physical factors
simultaneously: misalignment, illumination variations, and corrupted pixels.
As we discuss further below, illumination variation can be dealt by
expressing the test image as a linear combination of an appropriate set of
training images. Similar representations have been exploited in
illumination-robust tracking (e.g., \cite{Belhumeur1999-PAA,Murase1995-IJCV}).
For robustness to gross errors, the $\ell^1$-norm of the residual is a more
appropriate objective function than the classical $\ell^2$-norm. Its use here
is loosely motivated by theoretical results due to Cand\`{e}s and Tao
\cite{CandesE2005-IT} (see also \cite{Wright2008-IT}). These two observations
lead us to pose the registration problem as the search for a set of
transformations and illumination coefficients that minimize the $\ell^1$-norm
of the representation error. We solve this problem using a generalized
Gauss-Newton method which solves a sequence of affine-constrained $\ell^1$-norm
minimization problems \cite{Osborne1990-JAMSSB,Jittorntrum1980-NM}. Each of
these problems can also be solved efficiently using recently developed
first-order techniques for $\ell^1$-minimization, which are reviewed in
\cite{YangA2010-pp}.

@@

\subsection{Contributions} In this paper, we show how the 
mutually compounding issues of registration and illumination
can be naturally addressed within a robust sparse representation
framework. We show that face registration, a challenging
nonlinear problem, can be solved by a series of linear programs
that iteratively minimize the sparsity of the registration
error. This leads to an efficient and effective alignment
algorithm for face images that works for a large range of
variation in translation, rotation, and scale, even when
the face is only partially visible due to eyeglasses, closed
eyes and open mouth, sensor saturation, etc.  We also propose a
sufficient set of training illuminations
that is capable of interpolating typical indoor and outdoor
lighting, along with a practical hardware system for capturing
them.

@@

Finally, we demonstrate the effectiveness of the proposed new
methods with a complete face recognition system that is {\em
simple, stable, and scalable}. The proposed system performs
robust automatic recognition of subjects from loosely
controlled input images taken both indoors and outdoors, using
labeled frontal views of the subjects' faces under the proposed
illuminations for training and an off-the-shelf face
detector\footnote{We use the OpenCV
implementation of the Viola and Jones' face detector
\cite{Viola2004-IJCV}.} to detect faces in images.

@@

\noindent{\em Organization of this paper:} In Section
\ref{sec:registration}, we derive our robust registration and
recognition algorithm within the sparse representation
framework. We elaborate on implementation issues with our algorithm, conduct region of attraction experiments with
respect to both 2D in-plane deformation and 3D pose variation,
and discuss its relationship to existing work. Section \ref{sec:illumination} is dedicated to our
training acquisition system. Using this system, we investigate empirically the set of training illuminations required to handle practical illumination variations, and suggest a sufficient set of 38 training illuminations. Extensive experiments on
a large-scale public database and on our own database are conducted
in Section \ref{sec:multipie} and Section \ref{sec:own-data},
respectively, to verify the proposed system. Section
\ref{sec:conclusion} concludes our work with discussion of
promising future directions.

@@

Meanwhile, although our algorithm is not designed for
recognition when there is only a single gallery image per user,
and will not achieve competitive performance, we compare its
performance with LBP within such setting for the purpose of
better evaluation of the strength and limitation of both
algorithms. For this experiment, we use the FERET dataset
\cite{phillips1998feret}, which contains five standard
partitions: `fa' is the gallery containing 1196 frontal images
of 1196 subjects, and `fb', `fc', `dup1' and `dup2' are four
sets of probe images. The testing sets differ from the training
in facial expression (`fb'), illumination (`fc'), aging (`dup1'
) and long aging (`dup2'). In fact, except for `fb', we notice
significant changes of illumination in all the other three test
sets. For the training, we again crop and normalize the face
region from each original image to an $80\times 60$ window
using manually marked eye coordinates \cite{Deng2010-PR}. In
Table~\ref{tab:FERET-recognition}, we report the performance of
our algorithm on the four test sets, with input directly
obtained from the Viola and Jones' detector, and the
performance of LBP with input obtained from both face detector
and manual alignment using the eye coordinates.

@@

All these experimental results re-emphasize the importance that
both illumination and alignment need to be simultaneously handled
well in order to obtain accurate face recognition, even when there is
no obvious occlusion or corruption in the test.

@@

\section{Conclusion}\label{sec:conclusion}
Using a well-though-out combination of existing ideas
(iterative image alignment, l1-error function, SRC, using projectors for
illumination), one can build a system for recognizing human faces
from images taken under practical conditions that is conceptually simple, well
motivated, and competitive with state-of-the-art recognition systems for access
control scenarios.

The system achieves extremely {\em stable} performance under
a wide range of variations in illumination, misalignment, and even under small amounts of
pose and occlusion. We achieve very good recognition performance on
large-scale tests with public datasets as well as our practical face
images, while using only frontal 2D images in the training without
any explicit 3D face model.
Our system could potentially be extended to better handle large pose
and expression, either by incorporating training images with different poses or
expressions or by explicitly modeling and compensating the associated deformations
in the alignment stage.

>>>>

\PARstart{F}{ace} recognition applications to date have fallen into roughly two
categories.  Face recognition has recently seen a lot of success in a family of
less-demanding applications such as online image search and family photo album
organization (e.g.\ Google Picassa, Microsoft Photo Gallery, and Apple iPhoto).
At the other end of the tractability spectrum there are the terrorist watchlist
and mass surveillance applications that have for the most part dominated the
field of face recognition research.  However, there are many face recognition
that fall roughly between these extremes, where very high recognition
performance is desired, but the users in the gallery are still allies of the
system rather than adversaries.  These applications include access control for
secure facilities (e.g., prisons, office buildings), computer systems,
automobiles, or automatic teller machines, where controlled gallery images can
be obtained in advance.  These applications are very interesting due to their
potential sociological impact.  Since the gallery subjects are allies, rather
than opponents, of the recognition system, this creates the possibility of
carefully controlling the acquisition of the training data. While the same can
be said for other biometrics such as fingerprints and iris recognition, face
recognition has the potential of working with test data that is much less
controlled, allowing the access control system to be made less intrusive to the
users of the system. To some extent, the goal of this paper is to show how a
reliable face recognition system can be built for this restricted,
but still important, scenario.

@@

Unfortunately, while the work \cite{Wright2009-PAMI} achieves impressive results, it does not deal with misalignment
between the test and training images, and it requires a rich set of illuminations
in the gallery images for good performance.  
%Furthermore, the experiments were carried out on public datasets taken under
%controlled laboratory conditions such as Extended Yale B
%\cite{Georghiades2001-PAMI}, which do not contain realistic test images.
%two critical aspects of real world face recognition must be addressed:  significant variations in both the {\em image domain} and in
%the {\em image value}.
We illustrate the compounded effect of both alignment and illumination 
with an example in Figure \ref{fig:promo}.
The task is to identify the girl among 20 subjects. If the test
face image, say obtained from an off-the-shelf face detector,
has even a small amount of registration error against the
training images (caused by mild pose, scale, or misalignment),
the sparse representation obtained using the method of \cite{Wright2009-PAMI} is no longer informative, 
even if sufficient illuminations are present in the training, as shown in Figure
\ref{fig:promo}(top). Additionally, in order to span
the illuminations of a typical indoor (or outdoor) environment,
illuminations from behind the subject are needed in the
training set. Otherwise, even for perfectly aligned test images,
the sparse representation obtained using \cite{Wright2009-PAMI} will not necessarily be sparse or
informative, as shown by the example in Figure
\ref{fig:promo}(middle). Clearly, both good alignment, as well
as sufficient training images are needed to ensure success of the sparsity-based 
recognition method proposed by \cite{Wright2009-PAMI}.  
%Unfortunately, most public face databases lack images with a
%significant component of rear (more than 90 degrees from
%frontal) illumination, either for training or testing.
In this paper, we examine how to handle alignment and illumination simultaneously
in the sparse representation framework, bringing the method proposed in \cite{Wright2009-PAMI}
closer to practical use.

@@

We briefly review existing techniques for recognition, image registration, and
handling of illumination variation.  Our system is based purely on 2D
techniques.  This fact immediately distinguishes our approach from systems
that either require a 3D data as an input, or attempt to estimate a 3D model
from 2D input \cite{amberg2007reconstructing,Blanz2003-PAMI}.  While these
techniques can achieve better robustness to pose variation given
a sufficiently accurate 3D model, for access control applications where only 
moderate pose variation is present, the proposed method will be more
than sufficient.   Note that 2D images of faces
under varying illuminations already contain 3D shape-related information,\footnote{In principle,
one can recover the 3D shape of the face from multiple illuminations using photometric stereo.} 
and this information can be leveraged by 2D algorithms for alignment and recognition 
even if shape is not reconstructed explicitly. 
%In addition, the proposed sparsity-based robust
%alignment and recognition techniques are not limited to 2D data and can be easily extended
%to 3D data if necessary.

% Alignment
In holistic recognition algorithms, correspondence between points in the test
image and in the training must be achieved.  A long line of research exists on
using Active Appearance Models \cite{Cootes2001-PAMI}, and the closely related
Active Shape Models \cite{cootes1992active} to register images against a
relatively high-dimensional model of plausible face appearances, often
leveraging face-specific contours.  While these model-based techniques have
advantages in dealing with variations in expression and pose, they may add
unnecessary complexity to applications where subjects normally present a neutral face
or only have moderate expression. We prefer to focus on deformations with far fewer 
degrees of freedom, i.e. similarity transformations, and to use the training images themselves 
as the appearance model. Iterative registration in this spirit dates at least back to the 
Lucas-Kanade algorithm \cite{lucas1981iterative}.

Whereas much of the early work on image registration is aimed at the
problem of registering nearly identical images, say by minimizing a sum of
squared distances or maximizing normalized correlation, here we must confront several physical factors
simultaneously: misalignment, illumination variations, and corrupted pixels.
As we discuss further below, illumination variation can be dealt by
expressing the test image as a linear combination of an appropriate set of
training images. Similar representations have been exploited in
illumination-robust tracking (e.g., \cite{Belhumeur1999-PAA,Murase1995-IJCV}).
For robustness to gross errors, the $\ell^1$-norm of the residual is a more
appropriate objective function than the classical $\ell^2$-norm. Its use here
is loosely motivated by theoretical results due to Cand\`{e}s and Tao
\cite{CandesE2005-IT} (see also \cite{Wright2008-IT}). These two observations
lead us to pose the registration problem as the search for a set of
transformations and illumination coefficients that minimize the $\ell^1$-norm
of the representation error. We solve this problem using a generalized
Gauss-Newton method which solves a sequence of affine-constrained $\ell^1$-norm
minimization problems \cite{Osborne1990-JAMSSB,Jittorntrum1980-NM}. Each of
these problems can also be solved efficiently using recently developed
first-order techniques for $\ell^1$-minimization, which are reviewed in
\cite{YangA2010-pp}.

@@

\subsection{Contributions} In this paper, we show how 
registration and illumination
can be simultaneously addressed within a robust sparse representation
framework. We show that face registration, a challenging
nonlinear problem, can be solved by a series of linear programs
that iteratively minimize the sparsity of the registration
error. This leads to an efficient and effective alignment
algorithm for face images that works for a large range of
variation in translation, rotation, and scale, even when
the face is only partially visible due to eyeglasses, closed
eyes and open mouth, sensor saturation, etc.  We also propose a
sufficient set of training illuminations
that is capable of linearly representing typical indoor and outdoor
lighting, along with a practical hardware system for capturing
them.

@@

We then demonstrate the effectiveness of the proposed new
methods with a complete face recognition system that is {\em
simple, stable, and scalable}. The proposed system performs
robust automatic recognition of subjects from loosely
controlled probe images taken both indoors and outdoors, 
using a gallery of 
frontal views of the subjects' faces under the proposed
illuminations. An off-the-shelf face
detector\footnote{We use the OpenCV
implementation of the Viola and Jones' face detector
\cite{Viola2004-IJCV}.} is used to detect faces in the test images.

@@

\noindent{\em Organization of this paper:} In Section \ref{sec:registration},
we derive our robust registration and recognition algorithm within the sparse
representation framework. We elaborate on algorithmic implementation issues,
conduct region of attraction experiments with respect to both 2D in-plane
deformation and 3D pose variation, and discuss its relationship to existing
work. Section \ref{sec:illumination} is dedicated to our training acquisition
system. Using this system, we investigate empirically how many training
illuminations are required to handle practical illumination variations, and
suggest a sufficient set of 38 training illuminations. Extensive experiments on
a large-scale public database and on our own database are conducted in Section
\ref{sec:multipie} and Section \ref{sec:own-data}, respectively, to verify the
proposed system. Section \ref{sec:conclusion} concludes our work with
discussion of promising future directions.

@@

Although our algorithm is not designed for
recognition when there is only a single gallery image per user,
we compare its
performance with LBP within this setting for completeness.
For this experiment, we use the FERET dataset
\cite{phillips1998feret}, which contains five standard
partitions: `fa' is the gallery containing 1196 frontal images
of 1196 subjects, and `fb', `fc', `dup1' and `dup2' are four
sets of probe images. The testing sets differ from the training
in facial expression (`fb'), illumination (`fc'), aging (`dup1'
) and long aging (`dup2'). In fact, except for `fb', we notice
significant changes of illumination in all the other three test
sets. For the training, we again crop and normalize the face
region from each original image to an $80\times 60$ window
using manually marked eye coordinates \cite{Deng2010-PR}. In
Table~\ref{tab:FERET-recognition}, we report the performance of
our algorithm on the four test sets, with input directly
obtained from the Viola and Jones' detector, and the
performance of LBP with input obtained from both face detector
and manual alignment using the eye coordinates.

@@

All of these experimental results confirm that
both illumination and alignment need to be simultaneously handled
well in order to achieve accurate face recognition, even when there is
no obvious occlusion or corruption in the test.

@@

\section{Conclusion}\label{sec:conclusion}
Using a well-though-out combination of existing ideas
(iterative image alignment, l1-error function, SRC, using projectors for
illumination), we have proposed a system for recognizing human faces
from images taken under practical conditions that is conceptually simple, well
motivated, and competitive with state-of-the-art recognition systems for access
control scenarios.

The system achieves extremely stable performance under
a wide range of variations in illumination, misalignment, and even under small amounts of
pose and occlusion. We achieve very good recognition performance on
large-scale tests with public datasets as well as our practical face
images, while using only frontal 2D images in the gallery and no
explicit 3D face model.
Our system could potentially be extended to better handle large pose
and expression, either by incorporating training images with different poses or
expressions or by explicitly modeling and compensating the associated deformations
in the alignment stage.
